Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-26 draft.tex
Rewrote Image Data section

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{Centerline}
  \label{fig:centerline}
\end{figure}
\com{From VIS'14 Paper}
The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32]. This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra. Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation. The model uses a weighted sum of T1and T2-weighted MR images to detect the lumbar spine shape. Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21]. Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects. We obtained and worked with 2,540 tetrahedron models of the lumbar spine. For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
%%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{KlemmVMV13, KlemmVIS14}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline} (b)):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\noindent Spine shape is confounded by several somatometric variables.
%%
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
%%
Since men are on average taller than women, gender is another confounder.
%%
Large body weight increases the spine load, resulting in a bend shape.
%%
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
%%
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
%%
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{KlemmVIS14}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{%%
A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
%%
Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
%%
The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
%%
No correlation with back pain are present can be seen in this plot.
}
  \label{fig:image-parameter-range}
\end{figure*}
%%
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
%%
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
%%
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
%%
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
%%
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
%%
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
%%
The projections to the transversal planes attract attention as they have many outliers.
%%
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
%%
The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
%%
Curvature, curvature angle and torsion correlate strongly with their planar projections.
%%
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%%
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
%%
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
%%
No statistical significant correlation could be observed through all subject groups.
%%
The box plots show no difference between subjects with and without back pain.

@@

As described in a prior work \cite{Klemm2014}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.

@@

In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
%%
As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014}.
%%

>>>>

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{Centerline}
  \label{fig:centerline}
\end{figure}
\com{Image Data}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model (FEM) is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline} (b)):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\noindent Spine shape is confounded by several somatometric variables.
%%
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
%%
Since men are on average taller than women, gender is another confounder.
%%
Large body weight increases the spine load, resulting in a bend shape.
%%
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
%%
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
%%
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{Klemm2014VIS}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{%%
A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
%%
Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
%%
The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
%%
No correlation with back pain are present can be seen in this plot.
}
  \label{fig:image-parameter-range}
\end{figure*}
%%
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
%%
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
%%
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
%%
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
%%
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
%%
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
%%
The projections to the transversal planes attract attention as they have many outliers.
%%
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
%%
The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
%%
Curvature, curvature angle and torsion correlate strongly with their planar projections.
%%
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%%
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
%%
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
%%
No statistical significant correlation could be observed through all subject groups.
%%
The box plots show no difference between subjects with and without back pain.

@@

As described in a prior work \cite{Klemm2014VIS}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.

@@

In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
%%
As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
%%
